``Interestingness'' in Mathematical Theory Exploration

### 2019-10-25 PhDThesis/bucketing.tex
Wording
Bucketing content

Despite the clever search strategies employed by theory exploration tools, they
all rely on enumerating combinations of definitions in some way, which can cause
infeasible running times for input signatures with more than around a dozen
definitions (based on experiments with our Theory Exploration Benchmark, in
\S~\ref{sec:benchmark}. To avoid this, users of these systems must carefully
select only small subsets of their definitions to explore at a time.

We dub such cherry-picking the \emph{signature selection problem} for theory
exploration, and we analyse its effect on the running time of theory exploration
tools on the Theory Exploration Benchmark, and on the quality of the statements
they are able to generate. An automated solution to this signature selection
problem would accept a large number of definitions and efficiently select
sub-sets which are both small enough to reasonably explore, whilst still
enabling interesting properties to be discovered.

@@

\begin{enumerate}
\item Define the \emph{signature selection problem} for theory exploration.
\item Analyse the effect of signature selection on the quantity of desirable
  properties attainable by theory exploration on the Theory Exploration
  Benchmark problem set.
\item Identify the effect of ``toxic'' definitions on system performance, as
  measured by the Theory Exploration Benchmark.
\item Demonstrate how the signature selection problem is distinct from the
  well-known \emph{clustering problem} in machine learning, and how clustering
  algorithms cannot be directly applied to signature selection.
\item Propose recommendations to take signature selection and toxic definitions
  into account.
\end{enumerate}

@@

    \textbf{Right} Proportion of each population which timed out after 300
    seconds, for signatures of different size. This corresponds to a slice
    through the corresponding survival plot at the 300 second mark (if all of
    the measured populations were included). Lines of best-fit use linear
    regression, providing an estimate of the effect that ``toxic'' definitions
    have on failure (timeout) rate.}
    % TODO Put line slopes and error on graph
    % TODO Bottom row should use a second set of samples (chosen to avoid toxic
    % definitions), rather than just filtering down
  \label{fig:survival}
\end{figure}

@@

\iffalse
\begin{figure}
  \scalebox{0.45}{\input{images/timeoutsall.pgf}}
  \scalebox{0.45}{\input{images/timeoutsnontoxic.pgf}}
  \caption{Proportion of samples which timed out per size, with least-squares
    linear regression. First plot is for all TIP definitions, second removes
    runs given ``toxic'' definitions.}
  \label{fig:tailsize}
\end{figure}
\fi

@@

\begin{figure}
  \scalebox{0.45}{\input{images/proportionsall.pgf}}
  \scalebox{0.45}{\input{images/proportionsnontoxic.pgf}}
  \caption{Definitions, ordered by the ratio of successes to failures of the
    runs they appeared in. The first graph contains all TIP definitions, showing
    ``toxic'' definitions which always failed. The second graph only contains
    runs without any toxic definitions.
    \iffalse
    TODO: Alison: ? (ME: The green/red bar graphs appear as empty white boxes in
    the PDF Alison sent over. Check how it's rendered!)
    TODO: The x-axis ticks are crap (all overlapping); turn them off, put a
    single label like ``Definitions (ordered by failure rate)''
    \fi}
  \label{fig:proportions}
\end{figure}

@@

\begin{itemize}
\item Avoid ``toxic'' definitions: those producing exponentially large outputs,
  and/or taking many arguments (leading to a large number of combinations)
  \iffalse TODO: Alison: Is it always obvious which are toxic? \fi. In
  particular, tail-recursive functions which expose their accumulator argument
  are harder to explore. \iffalse TODO: Alison: These should be in the
  contributions and further recommendations based on your other findings in this
  chapter \fi
\item Smaller signatures are faster to explore, so should be preferred if
  possible.
\end{itemize}

It also seems prudent for theory exploration tools to limit time and memory
usage, both at the global level (to abort if it becomes clear we are in the
``long tail'') and at the level of individual evaluations. In particular such
fine-grained limits can abort evaluation of toxic definitions, whilst still
being lenient enough to allow the majority of non-toxic definitions to be
explored unhindered.

@@

Our final criterion \iffalse TODO: Alison: This looks odd since you haven't
mentioned criteria before \fi concerns the ``quality'' of the selected
subsets. This depends crucially on how we define whether a property is
``interesting''. We can use the same ground truth definition as our Theory
Exploration Benchmark: a property is ``interesting'' if it appears in a
particular corpus (in our case TIP~\cite{claessen2015tip}), and
``uninteresting'' if not. Whilst na\"ive, this approach allows the criterion to
be quantified: the quality of the selected subsets is the proportion of the
whole signature's interesting properties that also apply to the subsets taken
individually.

Since signature selection is largely independent of which tool we choose to
explore the resulting subsets, we do not require that such exploration actually
\emph{finds} all of the interesting properties; only that we have not
\emph{prevented} them from being found. For example, we may be given a signature
containing addition and multiplication functions and may consider commutativity
and distributivity properties to be ``interesting''. A subset containing both
would be a high quality selection, since all of these properties are
\emph{available} to find, even if our chosen tool doesn't spot them. In
contrast, splitting addition and multiplication into \emph{separate} subsets is
lower quality, since this stops \emph{any} tool from finding the distributivity
property (since it involves both definitions).

To understand the quality impact of limiting ourselves to such subsets, we
applied four signature selection methods to signatures sampled from the Theory
Exploration Benchmark. Signatures varied in size from a single definition up to
100; for each size, we sampled 100 signatures.

All four methods were forced to place each definition in exactly one subset, to
focus on the effect of splitting apart definitions rather than e.g. duplicating
definitions across several subsets or dropping some entirely. The tested methods
are:

@@

Figure~\ref{fig:bounds} shows the impact of selection (specifically, dividing
definitions between subsets, with no skipping or duplication) on the proportion
of interesting properties that remain available for subsequent theory
exploration tools to discover. The shaded region shows the difference between
(average) quality, with optimal and pessimal selections starting out equal (when
dividing signatures of $N$ definitions into $N$ subsets, where there is no
choice other than putting each definition on its own) but quickly diverging as
choices become available. Note that the optimal and pessimal methods are
constrained to avoid empty or unequally-distributed sets, so they cannot divide
up a signature of size $N$ into more than $N$ subsets. Pseudorandom selection is
shown as a grey line, and although the cardinality of its subsets is not
constrained, it nevertheless appears consistently between the bounds.

>>>>

Despite the clever search strategies employed by theory exploration tools, they
all rely on enumerating combinations of definitions in some way, which can cause
infeasible running times for input signatures with more than around a dozen
definitions (based on experiments with our Theory Exploration Benchmark, in
\S~\ref{sec:benchmark}). To avoid this, users of these systems must carefully
select only small subsets of their definitions to explore at a time.

We dub such cherry-picking the \emph{signature selection problem} for theory
exploration, and we analyse its effect on the running time of theory exploration
tools on our Theory Exploration Benchmark (TEB), and on the quality of the
statements they are able to generate. An automated solution to this signature
selection problem would accept a large number of definitions and efficiently
select sub-sets which are both small enough to reasonably explore, whilst still
enabling interesting properties to be discovered.

@@

\begin{enumerate}
\item Define the \emph{signature selection problem} for theory exploration.
\item Analyse the effect of signature selection on the quantity of desirable
  properties attainable by theory exploration on the TEB problem set.
\item Identify the effect of ``toxic'' definitions on system performance, as
  measured by the TEB.
\item Recommend improvements for theory exploration systems, to avoid or
  mitigate the problem of toxic definitions.
\item Demonstrate how the signature selection problem is distinct from the
  well-known \emph{clustering problem} in machine learning, and how clustering
  algorithms cannot be directly applied to signature selection.
\end{enumerate}

@@

    \textbf{Right} Proportion of each population which timed out after 300
    seconds, for signatures of different size. This corresponds to a slice
    through the corresponding survival plot at the 300 second mark (if all of
    the measured populations were included). Lines of best-fit use least-squares
    linear regression, indicating the effect that ``toxic'' definitions have on
    failure (timeout) rate.}
    % TODO Put line slopes and error on graph
    % TODO Bottom row should use a second set of samples (chosen to avoid toxic
    % definitions), rather than just filtering down
  \label{fig:survival}
\end{figure}

@@

@@

\begin{figure}
  \scalebox{0.69}{\input{images/proportionsall.pgf}}
  \scalebox{0.69}{\input{images/proportionsnontoxic.pgf}}
  \caption[]{Each bar is a definition from our Theory Exploration Benchmark,
    showing the proportion of signatures they appeared in which were
    successfully explored (green) compared to those which timed out (red).
    Definitions are sorted from most successful to least. The left graph shows
    all \input{text/proportions_graph_count_all.tex} TEB definitions explored, with five ``toxic'' definitions which always
    failed. with s ratio of successes to failures of the runs they appeared in. The first graph contains all TIP definitions, showing
    ``toxic'' definitions which always failed. The second graph only contains
    runs without any toxic definitions.}
  \label{fig:proportions}
\end{figure}

@@

\begin{itemize}
\item Avoid, or de-prioritise, exploring definitions which appear ``toxic''.
  We have not yet identified an \emph{a priori} toxicity test, but contributing
  factors include the use of large/infinite types (such as numbers and lists),
  recursion across multiple arguments and matching input/output types (i.e.
  \emph{endomorphisms}, which can be combined in exponentially-many ways).
  In particular, exposing the accumulator argument of a tail-recursive function
  makes it harder to explore. \iffalse TODO: Alison: These should be in the
  contributions and further recommendations based on your other findings in this
  chapter \fi
\item Smaller signatures are faster to explore, so should be preferred if
  possible.
\end{itemize}

We also recommend that theory exploration systems allow limits to be placed on
their time and memory usage: both at the global level, to abort if it becomes
clear we are in the ``long tail'' (e.g. if the discovery rate approaches zero);
and at the level of individual evaluations. The latter is more invasive, but
provides robustness against the presence of toxic definitions: their evaluation
will still drain resources for no benefit, but other (non-toxic) definitions
will continue to be explored, which would not be the case with a global abort.

More elaborate alternatives could also be pursued, like dynamic scheduling to
prioritise exploring definitions with the highest discovery rate.

@@

Regardless of how fast a signature is to explore, we must also be concerned with
the ``quality'' of the output when selecting its contents. This depends
crucially on how we define whether a property is ``interesting''. We can use the
same ground truth definition as our Theory Exploration Benchmark: a property is
``interesting'' if it appears in a particular corpus (in our case
TIP~\cite{claessen2015tip}), and ``uninteresting'' if not. Whilst na\"ive, this
approach is quantified and measurable: the quality of the subset of definitions
chosen to explore is the proportion of the
whole signature's interesting properties that also apply to the subsets taken
individually.

Since signature selection is largely independent of which tool we choose to
explore the resulting subsets, we do not require that such exploration actually
\emph{finds} all of the interesting properties; only that we have not
\emph{prevented} them from being found. For example, we may be given a signature
containing addition and multiplication functions and may consider commutativity
and distributivity properties to be ``interesting''. A subset containing both
would be a high quality selection, since all of these properties are
\emph{available} to find, even if our chosen tool doesn't spot them. In
contrast, splitting addition and multiplication into \emph{separate} subsets is
lower quality, as this prevents \emph{any} tool finding the distributivity
property (since it involves both definitions).

To understand the quality impact of limiting ourselves to such subsets, we
devised four signature selection methods and applied them to signatures sampled
from TEB. We varied the size of signatures from a single definition up to 100;
for each size, we sampled 100 signatures.

All four methods were forced to place each definition in exactly one subset, to
focus on the effect of splitting apart definitions rather than e.g. duplicating
definitions across several subsets or dropping some entirely (practical
implementations will not have these constraints, but it is important here to
avoid gaming our evaluation). The tested methods are:

@@

Figure~\ref{fig:bounds} shows the impact of selection (specifically, dividing
definitions between subsets, with no skipping or duplication) on the proportion
of interesting properties that remain available for subsequent theory
exploration tools to discover. The shaded region shows the difference between
the average quality of optimal and pessimal methods. These always start equal
(when dividing signatures of $N$ definitions into $N$ subsets, which can only be
done in one way) but quickly diverging as choices become available. The average
quality of pseudorandom selection is shown as a grey line, and although the
cardinality of its subsets is not constrained, it nevertheless appears
consistently between the bounds.
